---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 10 @ 11:59PM
author: YOUR NAME and UID
format:
  html:
    theme: cosmo
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    cache: false    
    echo: true
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
---

---
Display machine information for reproducibility:
```{r}
#| eval: false
sessionInfo()
```

Load necessary libraries (you can add more as needed).
```{r setup}
library(data.table)
library(lubridate)
library(R.utils)
library(tidyverse)
```

MIMIC data location
```{r}
mimic_path <- "~/mimic"
```

In this exercise, we use tidyverse (ggplot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2023winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

Display the contents of MIMIC data folder. 
```{r}
system(str_c("ls -l ", mimic_path, "/"), intern = TRUE)
system(str_c("ls -l ", mimic_path, "/core"), intern = TRUE)
system(str_c("ls -l ", mimic_path, "/hosp"), intern = TRUE)
system(str_c("ls -l ", mimic_path, "/icu"), intern = TRUE)
```
---

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. 

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

For later questions, we stick to the `read_csv` in tidyverse.

- 'read.csv'
```{r}
system.time(data1<-read.csv("/Users/huxuyi/mimic/core/admissions.csv.gz"))
```

-'read_csv'
```{r}
system.time(data2<-read_csv("/Users/huxuyi/mimic/core/admissions.csv.gz"))
```

-'fread'
```{r}
system.time(data3<-fread("/Users/huxuyi/mimic/core/admissions.csv.gz"))
```

*So, according to the results, 'fread' is the fastest.*

```{r}
str(data1)
```
```{r}
str(data2)
```
```{r}
str(data3)
```

*So, read.csv can not read datetime but read_csv and fread can.*


## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{r}
system(
  str_c(
    "gzcat < ", 
    str_c(mimic_path, "/icu/icustays.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `icustatys.csv.gz` as a tibble `icustays_tble`. 

```{r}
icustays_tble <- as_tibble(read_csv(str_c(mimic_path, "/icu/icustays.csv.gz")))
```


2. How many unique `subject_id`? Can a `subject_id` have multiple ICU stays? 

```{r}
n_distinct(icustays_tble$subject_id)
n_distinct(icustays_tble$stay_id)
```
53150 unique subject_id. Yes, since there are more stay_ids than subject_ids.

3. Summarize the number of ICU stays per `subject_id` by graphs. 

```{r}
freq<-icustays_tble[,c(1,3)] %>%
  group_by(subject_id) %>%
  summarise(stays=n())

ggplot(data=freq) + 
  geom_freqpoly(mapping = aes(x=stays)) +
  labs(title="Freqrency plot of the number of ICU stays per 'subject_id'")
```


4. For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`. (Hint: `slice_min` and `slice_max` may take long. Think alternative ways to achieve the same function.)

```{r}
first_icu<-icustays_tble[,c(1,3)] %>%
  group_by(subject_id) %>%
  slice_head(n=1)

slice_head(first_icu,n=10)
```


## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/hosp/admissions/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/core/admissions.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `admissions.csv.gz` as a tibble `admissions_tble`.
```{r}
admissions_tble<-as_tibble(read_csv(str_c(mimic_path,'/core/admissions.csv.gz')))
```

2. Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadmi_id`.

```{r}
slice_head(semi_join(admissions_tble,icustays_tble,key=c("subject_id"="subject_id","hadmi_id"="hadmi_id")),n=10)
```


3. Summarize the following variables by graphics. 

    - admission year  
    - admission month  
    - admission month day  
    - admission week day  
    - admission hour (anything unusual?)  
    - admission minute (anything unusual?)  
    - length of hospital stay (anything unusual?)    

```{r}
adm_time<-bind_cols(
  adm_year=year(admittime),
  adm_month=month(admittime),
  admi_md=mday(admittime),
  admi_wd=wday(admittime),
  admi_h=hour(admittime),
  admi_ms=minute(admittime),
  stay=difftime(dischtime,admittime,units=c('days'))
)
```

- admission year:
```{r}
ggplot(data=adm_time)+
  geom_freqpoly(mapping=aes(x=adm_year))+
  labs(x="Admission Year",y="Count",
       title="The number of patients in the admission year")
```

-admission month:
```{r}
ggplot(data=adm_time)+
  geom_bar(mapping=aes(x=adm_month))+
  labs(x="Admission Month",y="Count",
       title="The number of patients in the each month")
```

-admission month day:
```{r}
ggplot(data=adm_time)+
  geom_freqpoly(mapping=aes(x=admi_md))+
  labs(x="Admission Month Day",y="Count",
       title="The number of patients in the admission year")
```

- admission week day:
```{r}
ggplot(data=adm_time)+
  geom_bar(mapping=aes(x=admi_wd)) +
  labs(x="Admission Week Day",y="Count",
       title="The number of patients in the each weekday")
```
- Admission hour:
```{r}
ggplot(data=adm_time)+
  geom_freqpoly(mapping=aes(x=admi_h))+
  labs(x="Admission Hour",y="Count",
       title="The number of patients in a day")
```

There are hours smaller than 0.

-admission minutes:
```{r}
ggplot(data=adm_time)+
  geom_freqpoly(mapping=aes(x=admi_ms))+
  labs(x="Admission Minutes",y="Count",
       title="The number of patients with unit minutes")
```

There are numbers that smaller than 0 and larger than 60.

-Length of hospital stay
```{r}
df1=filter(adm_time,stay<quantile(stay,0.99))
ggplot(data=df1)+
  geom_density(mapping=aes(x=stay))+
  labs(x="Length of hospital stay",y="Count",
       title="The distribution of length of hospital stay of patients")

df1=filter(adm_time,stay<quantile(stay,0.99))
ggplot(data=df1)+
  geom_density(mapping=aes(x=stay))+
  labs(x="Length of hospital stay",y="Count",
       title="The distribution of length of hospital stay of patients")

ggplot(data=df1)+
  geom_boxplot(mapping=aes(x=stay))+
  labs(x="Length of hospital stay",
       title="The distribution of length of hospital stay of patients")
```
There are some extremely large outliers.

## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/hosp/patients/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/core/patients.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).

2. Summarize variables `gender` and `anchor_age`, and explain any patterns you see.

## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/hosp/labevents.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```
`d_labitems.csv.gz` is the dictionary of lab measurements. 
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/hosp/d_labitems.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Find how many rows are in `labevents.csv.gz`.

2. We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), and glucose (50931). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`. 

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` option in the `read_csv` function to reduce the memory burden. It took my computer 5-10 minutes to ingest this file. If your computer really has trouble importing `labevents.csv.gz`, you can import from the reduced data file `labevents_filtered_itemid.csv.gz`.

3. Further restrict `labevents_tble` to the first lab measurement during the ICU stay. 

4. Summarize the lab measurements by appropriate numerics and graphics. 

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/chartevents.csv.gz"), 
    " | head"), 
  intern = TRUE
)
```
`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/d_items.csv.gz"), 
    " | head"), 
  intern = TRUE
)
```

1. We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

    Hint: `chartevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` option in the `read_csv` function to reduce the memory burden. It took my computer >15 minutes to ingest this file. If your computer really has trouble importing `chartevents.csv.gz`, you can import from the reduced data file `chartevents_filtered_itemid.csv.gz`.

2. Further restrict `chartevents_tble` to the first vital measurement during the ICU stay. 

3. Summarize these vital measurements by appropriate numerics and graphics. 

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are the first ICU stay of each unique adult (age at admission > 18) and columns contain at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- all variables in `patients.csv.gz`  
- first lab measurements during ICU stay  
- first vital measurements during ICU stay
- an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)

## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

- `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)

- `thirty_day_mort` vs first lab measurements

- `thirty_day_mort` vs first vital measurements

- `thirty_day_mort` vs first ICU unit
